### 2014-09-10 paper/paper.Rmd
updated code for shortages, and for family structure

```{r load_packages, echo = FALSE, message = FALSE, warning = FALSE, results='hide'}
library(dplyr)
library(stringr)
library(reshape2)
library(ggplot2)
library(lubridate)
library(ggvis)
library(maps)
library(ggmap)
library(htmltools)
library(rworldmap)
library (grid)    
library (scales)
library(doBy)
```

@@

*** We should decide on country names, to coordinate map info and the test data. This si some really ugly code to convert names to be the same between the two. Could be made cleaner, much cleaner!

@@

### Teachers, computers, and libraries at school make a difference in math scores

```{r shortages, echo = FALSE, message = FALSE, warning = FALSE}
school2012.sub = school2012[, c(1, 6, 34, 35, 40, 45, 46, 47, 48, 49, 50, 51)] #%>% select(name, SCHOOLID, SC11Q02, SC11Q03, SC14Q02, SC14Q07, SC14Q08, SC14Q09, SC14Q10, SC14Q11, SC14Q12, SC14Q13)
student2012.sub = student2012[, c(1, 6, 501, 541, 546, 634)] #%>% select(name, SCHOOLID, PV1MATH, PV1READ, PV1SCIE)

scores_by_school = summarise(group_by(student2012.sub, name, SCHOOLID), mmath = mean(PV1MATH), 
                                      mread = mean(PV1READ), msci = mean(PV1SCIE))

df = scores_by_school %>% left_join(school2012.sub)
df2 <- summarise(group_by(df, name, SC14Q02), mmath2 = mean(mmath))
qplot(SC14Q02, mmath2, data = df2, facets = ~name)
df3 <- summarise(group_by(df, name, SC14Q10), mmath2 = mean(mmath))
qplot(SC14Q10, mmath2, data = df3, facets = ~name)
```

### The gender gap in math is not universal but the reading gap is, in favor of girls

@@

### Starting school at an older age generated lower average math score.

@@

### Time spent out of school studying is important, but only up to a point

@@

### Parents matter
### subsection: Number of parents in the home
Looking at the number of parents in a household shows that student in two-parent household are on average performing better than student's not in a two-parent household. Students with just a mother at home do tend to perform better than students with only a father at home. Students with no parents at home are performing the worst, on average.

```{r par_athomebar, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
athome = studentLuke[,c("name", "SCHOOLID", "STIDSTD", "OECD", "PV1MATH", "PV1SCIE", "PV1READ","ST11Q01", "ST11Q02", "ST11Q03","ST11Q04", "ST11Q05", "ST11Q06")]
athome = athome %>% filter(!is.na(athome$ST11Q01) & !is.na(athome$ST11Q02))
athome$parents = NA
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "Yes", ]$parents = "Both"
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "No", ]$parents = "Mother"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "Yes", ]$parents = "Father"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "No", ]$parents = "Neither"
athome$name <- reorder(athome$name,athome$PV1MATH,mean)
pbarhome <- qplot(x = parents, data = athome) 
pbarhome

```{r par_athome, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
phome <- qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, facets = ~name, data = athome) + coord_flip() + ylab("Math") + xlab("Family Structure") 
phome
```

### subsection: Parent's Occupation Status
One of the variables the survey addressed was the occupation status of each of the parents. There were five possible categories: “Full-Time”, “Part-Time”, “Not working, but looking for a job”, “Other (e.g. home duties, retired)”, and no response. The distribution of the number of parents who selected each of the categories can be seen in the bar chart. The chart shows that father’s primarily have full-time jobs. However, there are large numbers of mothers who work full-time, part-time, or are stay-at-home moms. With women making up a larger portion of the work-force, it is interesting to look at how the parent’s occupation relates to the student’s performance in mathematics, reading, and science.

@@

studentLuke$name <- as.character(studentLuke$CNT)
# unique(anti_join(studentLuke.sub, world.polys)[1])
studentLuke$name[studentLuke$name=="United Arab Emirates"] <- "UAE"
studentLuke$name[studentLuke$name=="United Kingdom"] <- "UK"
studentLuke$name[studentLuke$name=="Serbia"] <- "Serbia"
studentLuke$name[studentLuke$name=="Korea"] <- "South Korea"
studentLuke$name[studentLuke$name=="Chinese Taipei"] <- "Taiwan"
studentLuke$name[studentLuke$name=="Slovak Republic"] <- "Slovakia"
studentLuke$name[studentLuke$name=="Russian Federation"] <- "Russia"
studentLuke$name[studentLuke$name=="Perm(Russian Federation)"] <- "Russia"
studentLuke$name[studentLuke$name=="Hong Kong-China"] <- "Hong Kong"
studentLuke$name[studentLuke$name=="China-Shanghai"] <- "China"
studentLuke$name[studentLuke$name=="Macau"] <- "China"
studentLuke$name[studentLuke$name=="Connecticut (USA)"] <- "USA"
studentLuke$name[studentLuke$name=="Florida (USA)"] <- "USA"
studentLuke$name[studentLuke$name=="Massachusetts (USA)"] <- "USA"
studentLuke$name[studentLuke$name=="United States of America"] <- "USA"
unique(studentLuke$name)
studentLuke$name <- factor(studentLuke$name)

a6 <- summaryBy(data=studentLuke, PV1MATH + ESCS ~ ST15Q01+ name,FUN=mean,rm.na=T)
a7 <- summaryBy(data=studentLuke, PV1MATH + ESCS ~ ST19Q01+ name,FUN=mean,rm.na=T)
colnames(a6) <- colnames(a7) <- c("Job.Status","Country","Math","ESCS")
levels(a6$Job.Status) <- levels(a7$Job.Status)  <- c("Full-Time","Part-Time","Unemployed","Other","NA")

a8 <- rbind(cbind(a6,Parent="Mother"),cbind(a7,Parent="Father"))
#a2 <- melt(a1)
ind <- which(a8$Job.Status=="Other"& a8$Parent=="Mother")
a8$Country <- factor(a8$Country,levels(a8$Country)[order(a8$Math[ind])])
p <- qplot(Country, Math, col=Job.Status, data = a8) +coord_flip() + facet_wrap(~Parent)
p
```

@@

### The weird and wacky republic of Albania

@@

### More TVs yield higher math scores in the developing world, but lower in the developed countries

>>>>

```{r load_packages, echo = FALSE, message = FALSE, warning = FALSE, results='hide'}
library(dplyr)
library(stringr)
library(reshape2)
library(ggplot2)
library(lubridate)
#library(ggvis)
library(maps)
library(ggmap)
library(htmltools)
library(rworldmap)
library(grid)    
library(scales)
library(doBy)
```

@@

@@

# 1. Teachers, computers, and libraries at school make a difference in math scores

```{r shortages, echo = FALSE, message = FALSE, warning = FALSE}
school2012.sub = school2012[, c("name", "SCHOOLID", "SC14Q02", "SC14Q07", "SC14Q10")]
student2012.sub = student2012[, c("name", "SCHOOLID", "PV1MATH", "PV1READ", "PV1SCIE")]

scores_by_school = student2012.sub %>% group_by(name, SCHOOLID) %>% 
  summarise(mmath = mean(PV1MATH), mread = mean(PV1READ), msci = mean(PV1SCIE))

df = scores_by_school %>% left_join(school2012.sub)
df$SC14Q02 <- factor(df$SC14Q02, labels=c(0,1,2,3))
df$SC14Q07 <- factor(df$SC14Q07, labels=c(0,1,2,3))
df$SC14Q10 <- factor(df$SC14Q10, labels=c(0,1,2,3))
df.mshort <- df[!is.na(df$SC14Q02),] %>% group_by(name, SC14Q02) %>% summarise(mmath2 = mean(mmath, na.rm=T))
#qplot(SC14Q02, mmath2, data = df.mshort, facets = ~name, 
#      xlab="Math teacher shortage (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
#      ylab="Math Score", geom=c("point","line"), group=name)
df.lshort <- df[!is.na(df$SC14Q10),] %>% group_by(name, SC14Q10) %>% summarise(mmath2 = mean(mmath, na.rm=T))
#qplot(SC14Q10, mmath2, data = df.lshort, facets = ~name, 
#      xlab="Library shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
#      ylab="Math Score", geom=c("point","line"), group=name)
df2 <- summarise(group_by(df[!is.na(df$SC14Q02),], name, SC14Q02), teachers = mean(mmath, na.rm=T))
colnames(df2)[2] <- "levels"
df3 <- summarise(group_by(df[!is.na(df$SC14Q07),], name, SC14Q07), computers = mean(mmath, na.rm=T))
colnames(df3)[2] <- "levels"
df4 <- summarise(group_by(df[!is.na(df$SC14Q10),], name, SC14Q10), libmaterials = mean(mmath, na.rm=T))
colnames(df4)[2] <- "levels"
df.all <- merge(df2, df3)
df.all <- merge(df.all, df4)
df.m <- melt(df.all, id=c("name", "levels"))
colnames(df.m)[4] <- "math"
qplot(levels, math, data = df.m, facets = ~name, 
      xlab="Shortages (0=None, 1=Very little, 2=To some extent, 3=A lot)", 
      ylab="Math Score", geom=c("point","line"), group=variable, colour=variable) +
      scale_colour_manual("Type", values=c("teachers"="#1B9E77", "computers"="#D95F02", "libmaterials"="#7570B3")) +
      theme(legend.position="bottom") 
#ggsave("shortages.pdf", width=14, height=14)
qplot(SC14Q02, data=df, geom="histogram", xlab="Teacher shortage", fill=I("#1B9E77"))
#ggsave("teacher-shortages.pdf", width=5, height=5)
qplot(SC14Q07, data=df, geom="histogram", xlab="Computer shortage", fill=I("#D95F02"))
#ggsave("computer-shortages.pdf", width=5, height=5)
qplot(SC14Q10, data=df, geom="histogram", xlab="Library materials shortage", fill=I("#7570B3"))
#ggsave("library-shortages.pdf", width=5, height=5)
```

# 2. The gender gap in math is not universal but the reading gap is, in favor of girls

@@

# 3. Starting school at an older age generated lower average math score.

@@

# 4. Time spent out of school studying is important, but only up to a point

@@

# 5. Parents matter
### a. Number of parents in the home
Looking at the number of parents in a household shows that student in two-parent household are on average performing better than student's not in a two-parent household. Students with just a mother at home do tend to perform better than students with only a father at home. Students with no parents at home are performing the worst, on average.

```{r par_athomebar, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
athome = student2012[,c("name", "SCHOOLID", "STIDSTD", "OECD", "PV1MATH", "PV1SCIE", "PV1READ","ST11Q01", "ST11Q02", "ST11Q03","ST11Q04", "ST11Q05", "ST11Q06")]
athome = athome %>% filter(!is.na(athome$ST11Q01) & !is.na(athome$ST11Q02))
athome$parents = NA
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "Yes", ]$parents = "Both"
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "No", ]$parents = "Mother"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "Yes", ]$parents = "Father"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "No", ]$parents = "Neither"
athome$parents <- factor(athome$parents, levels=c("Both", "Mother", "Father", "Neither"))
#athome$name <- reorder(athome$name,athome$PV1MATH,mean, decreasing=F)
pbarhome <- qplot(x = parents, data = athome, xlab="Parents in home") 
pbarhome
```

```{r par_athome, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
phome <- qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, data = athome) + ylab("Math") + facet_wrap(~name, scales="free_y") + scale_x_discrete("Family Structure", labels=c("Mother"="M", "Father"="F", "Both"="B", "Neither"="N")) 
phome
```

### b. Parent's Occupation Status
One of the variables the survey addressed was the occupation status of each of the parents. There were five possible categories: “Full-Time”, “Part-Time”, “Not working, but looking for a job”, “Other (e.g. home duties, retired)”, and no response. The distribution of the number of parents who selected each of the categories can be seen in the bar chart. The chart shows that father’s primarily have full-time jobs. However, there are large numbers of mothers who work full-time, part-time, or are stay-at-home moms. With women making up a larger portion of the work-force, it is interesting to look at how the parent’s occupation relates to the student’s performance in mathematics, reading, and science.

@@

#studentLuke$name <- as.character(studentLuke$CNT)
# unique(anti_join(studentLuke.sub, world.polys)[1])
#studentLuke$name[studentLuke$name=="United Arab Emirates"] <- "UAE"
#studentLuke$name[studentLuke$name=="United Kingdom"] <- "UK"
#studentLuke$name[studentLuke$name=="Serbia"] <- "Serbia"
#studentLuke$name[studentLuke$name=="Korea"] <- "South Korea"
#studentLuke$name[studentLuke$name=="Chinese Taipei"] <- "Taiwan"
#studentLuke$name[studentLuke$name=="Slovak Republic"] <- "Slovakia"
#studentLuke$name[studentLuke$name=="Russian Federation"] <- "Russia"
#studentLuke$name[studentLuke$name=="Perm(Russian Federation)"] <- "Russia"
#studentLuke$name[studentLuke$name=="Hong Kong-China"] <- "Hong Kong"
#studentLuke$name[studentLuke$name=="China-Shanghai"] <- "China"
#studentLuke$name[studentLuke$name=="Macau"] <- "China"
#studentLuke$name[studentLuke$name=="Connecticut (USA)"] <- "USA"
#studentLuke$name[studentLuke$name=="Florida (USA)"] <- "USA"
#studentLuke$name[studentLuke$name=="Massachusetts (USA)"] <- "USA"
#studentLuke$name[studentLuke$name=="United States of America"] <- "USA"
unique(studentLuke$name)
studentLuke$name <- factor(studentLuke$name)

a6 <- summaryBy(data=studentLuke, PV1MATH + ESCS ~ ST15Q01+ name,FUN=mean,rm.na=T)
a7 <- summaryBy(data=studentLuke, PV1MATH + ESCS ~ ST19Q01+ name,FUN=mean,rm.na=T)
colnames(a6) <- colnames(a7) <- c("Job.Status","name","Math","ESCS")
levels(a6$Job.Status) <- levels(a7$Job.Status)  <- c("Full-Time","Part-Time","Unemployed","Other","NA")

a8 <- rbind(cbind(a6,Parent="Mother"),cbind(a7,Parent="Father"))
#a2 <- melt(a1)
ind <- which(a8$Job.Status=="Full-Time"& a8$Parent=="Mother")
a8$name <- factor(a8$name,levels(a8$name)[order(a8$Math[ind])])
p <- qplot(name, Math, col=Job.Status, data = a8) +coord_flip() + facet_wrap(~Parent)
p
```

@@

# 6. Albania is not like anyone else!

@@

# 7. More TVs yield higher math scores in the developing world, but lower in the developed countries
